\section{OpenMP--MPI Roadmap}

The OpenMP--MPI backend evolved through successive design iterations. The initial hybrid baseline established a functional starting point, but its scaling behavior exposed several architectural bottlenecks. A later iteration redesigned the solver according to high-performance computing principles, targeting cache efficiency, reduced synchronization, and communication patterns that remain sustainable on large instances. This section summarizes that evolution and explains why a subsequent attempt at \emph{intra-ant parallelism} on CPUs failed despite its apparent theoretical appeal.

\subsection{Initial iteration: baseline limitations}

The initial hybrid implementation already exploited the natural mapping of ACO to distributed-memory execution: each MPI rank acted as a sub-colony, while OpenMP threads constructed local ant solutions. In practice, however, four bottlenecks limited both strong and weak scaling.

\begin{itemize}
    \item \textbf{False sharing.} Telemetry and per-thread statistics were stored in non-aligned arrays of structures. Frequent writes from different threads caused repeated cache-line invalidations in L1/L2, reducing effective core throughput.
    \item \textbf{Fork--join overhead.} The OpenMP parallel region was entered and destroyed once per epoch. This imposed a fixed latency at every iteration, which became visible especially on medium-size instances with short epochs.
    \item \textbf{Serialized pheromone deposit.} The deposit phase was executed inside a \texttt{\#pragma omp single} region, introducing a strict serial tail at the end of each epoch and therefore an Amdahl-style limit on speedup.
    \item \textbf{High MPI latency.} Pheromone synchronization was performed through one \texttt{MPI\_Iallreduce} per row of the matrix, that is, $N+1$ collectives per epoch. The resulting stream of small messages saturated the communication layer long before the network bandwidth itself became the limiting resource.
\end{itemize}

These issues meant that the baseline solver was parallel in structure but not yet HPC-ready in execution behavior.

\subsection{Version~2: Architectural pillars}

Version~2 restructured the OpenMP--MPI backend around four complementary ideas.

\subsubsection{Persistent threading}

The OpenMP region is created once outside the epoch loop and kept alive for the full run. Threads synchronize through lightweight barriers and shared state transitions instead of repeated fork--join cycles. This removes a fixed source of overhead and improves temporal locality in thread-private workspaces.

\subsubsection{Cache awareness and adaptive auto-tuning}

The second pillar is an explicitly cache-oriented data layout:

\begin{itemize}
    \item all thread-private structures are padded and aligned to 64~bytes, matching the cache-line size of the target CPUs;
    \item hot buffers are organized to avoid inter-thread interference;
    \item the parameter $K_{cand}$ is auto-tuned at startup from the detected L3 capacity.
\end{itemize}

The tuning rule was refined to account for both candidate indices and scores, for an effective footprint of 8~bytes per element with a 70\% safety factor. The goal is to keep the hottest exploration structures resident in L3 and therefore reduce fallback traffic toward DRAM.

\subsubsection{Fully parallel pheromone update}

The serialized deposit phase of Version~1 was replaced by a fully parallel update based on hardware atomics:

\begin{lstlisting}[language=C, caption=Parallel pheromone deposit with OpenMP atomics]
#pragma omp parallel for schedule(guided)
for (int ant = 0; ant < local_m; ++ant) {
    for (int e = 0; e < route_len[ant] - 1; ++e) {
        int u = route[ant][e];
        int v = route[ant][e + 1];
        double delta = deposit[ant][e];
        #pragma omp atomic
        tau[u][v] += delta;
    }
}
\end{lstlisting}

Although atomic operations introduce contention on hot edges, this design removes the dominating serial region and shifts the deposit phase back into the parallel fraction of the code.

\subsubsection{Sparse delta synchronization over MPI}

The most consequential redesign concerned communication. Instead of synchronizing the full dense pheromone matrix, each rank emits only the modified edges accumulated during the epoch. Communication therefore moves from a dense model to an incremental sparse-delta model based on index--increment pairs.

\begin{lstlisting}[language=C, caption=Sparse pheromone exchange through delta aggregation]
pack_local_deltas(local_idx, local_val, &local_nnz);
MPI_Allgatherv(local_idx, local_nnz, MPI_INT,
               all_idx, recvcounts, displs, MPI_INT,
               MPI_COMM_WORLD);
MPI_Allgatherv(local_val, local_nnz, MPI_DOUBLE,
               all_val, recvcounts, displs, MPI_DOUBLE,
               MPI_COMM_WORLD);
apply_remote_deltas(all_idx, all_val, total_nnz);
\end{lstlisting}

On the large instances considered in this report, this change reduces network traffic by roughly three orders of magnitude: from about 1~GB of dense state to about 2~MB of actual modifications.

\subsection{Empirical validation of Version~2}

The later experimental campaigns confirmed that the sparse design matches the true behavior of the solver.

\subsubsection{Experiment~15: sparsity analysis}

For $N=16{,}000$ and $M=1536$, the measured update density per epoch is only \textbf{0.1085\%}. In absolute terms, this means that on a 1~GB pheromone matrix only about \textbf{2.12~MB} of data are modified in a typical epoch. This result justifies abandoning dense collectives in favor of sparse communication.

\subsubsection{Experiment~16: incremental sparse MPI synchronization}

Replacing row-wise \texttt{MPI\_Iallreduce} with \texttt{MPI\_Allgatherv} over packed deltas produced a measured speedup of \textbf{1.61\texttimes} in a local 2-rank setup. The improvement came from eliminating the transfer of almost entirely zero-valued matrix regions. To avoid shifting the bottleneck elsewhere, pheromone normalization was also parallelized instead of being left to the master thread.

\subsubsection{Experiment~17: sparse asynchronous SSP}

The final Version~2 configuration combines sparse synchronization with \emph{Stale Synchronous Parallelism} (SSP) using \texttt{MPI\_Iallgatherv}. In the local shared-memory testbed the observed speedup was \textbf{1.03\texttimes}, while the expected benefit on a real multi-node cluster is estimated in the \textbf{1.2\texttimes--1.5\texttimes} range. The modest local gain is coherent with the extremely low latency of local MPI paths: once the network is already cheap, overlapping communication with computation removes only a small residual cost. On real clusters, however, the asynchronous design moves the network almost entirely out of the critical path.

\subsection{Current status and next steps}

The current solver can be summarized as an \textbf{adaptive, hierarchical, sparse-SSP} OpenMP--MPI engine. Within the scope of this project, its main properties are:

\begin{enumerate}[leftmargin=*,itemsep=2pt]
    \item \textbf{Network efficiency:} approximately a 460\texttimes{} reduction in MPI traffic, from dense gigabyte-scale synchronization to sparse megabyte-scale deltas.
    \item \textbf{Computation--communication overlap:} useful work and collective progress proceed concurrently rather than as two strictly separated phases.
    \item \textbf{High throughput on large instances:} about 770~ms per epoch at $N=16{,}000$, with near-ideal local scaling in the tested environment.
\end{enumerate}

The next logical steps are to validate the SSP design on 4--8 physical cluster nodes and to study a hybrid CPU--GPU extension in which the GPU accelerates score-matrix computation while the CPU retains sparse route construction and communication management.

\subsection{Why intra-ant parallelism failed on CPUs}

Even after the improvements above, the construction stage remained the dominant cost on very large instances, accounting for roughly 75\% of the runtime at $N=16{,}000$. A new Version~3 therefore explored \textbf{collaborative ant teams}: instead of mapping one ant to one thread, several threads cooperated on the same ant, with the goal of accelerating the fallback scan over distance rows of about 64~KB.

Three variants were implemented, and all of them failed.

\subsubsection{V3.1: synchronous team}

The first design used explicit OpenMP barriers between the leader thread, responsible for the probabilistic choice, and the worker threads, responsible for the fallback scan. With approximately $1536$ ants and $16{,}000$ construction steps per epoch, the implementation executed on the order of \textbf{24 million global barriers per epoch}. The measured epoch time rose to about \textbf{8.4~s}, compared with about \textbf{1.4~s} for the Version~2 design based on independent threads.

\subsubsection{V3.2: speculative spinning}

To remove barrier overhead, the second design replaced explicit barriers with generation-based spinning on shared atomic counters. In theory this avoided software barriers; in practice it created a \textbf{coherency storm}. Continuous polling on shared cache lines triggered MESI invalidations and cache-line bouncing, slowing the leader thread exactly when it needed bandwidth to read the real problem data. The execution became unstable and exceeded \textbf{5 minutes per epoch}.

\subsubsection{V3.3: on-demand wakeup}

The third design used \texttt{pthread\_cond\_wait} to park worker threads and wake them only when a heavy fallback scan was required. This eliminated busy-waiting but introduced a new bottleneck: kernel wakeup latency. A wakeup cost of about \textbf{5--10~$\mu$s} is already comparable to, or larger than, the useful serial scan itself. Consequently, cores looked underutilized not because they had no work, but because they were repeatedly waiting on the operating system scheduler.

\subsection{Technical interpretation of the failure}

The failure of Version~3 is fundamentally a mismatch between task granularity and coordination cost. The fallback over a 64~KB row is a micro-granular memory task. On a modern CPU, the serial work is so short that any inter-thread protocol becomes proportionally too expensive. The parallel execution time can be expressed as

\[
T_{par} = \frac{T_{comp}}{P} + T_{coord},
\]

where $T_{comp}$ is the serial scan time, $P$ is the number of collaborating workers, and $T_{coord}$ is the cost of synchronization. In our tests,

\begin{itemize}
    \item $T_{comp} \approx 10~\mu s$,
    \item $T_{coord} \approx 15$--$20~\mu s$.
\end{itemize}

Since $T_{coord} > T_{comp}$, the collaborative version is slower than the serial one for any practical value of $P$. The issue is not a poor implementation of a promising idea; rather, it is a limit imposed by the physical cost of inter-core coordination on commodity multicore CPUs.

\subsection{Design lesson}

Taken together, these results point to a clear design principle for CPU-based ACO solvers: \textbf{one ant per thread} remains the most robust strategy for sustaining high throughput. If the goal is instead to reduce the latency of the fallback stage, the more promising path is not intra-ant collaboration across cores, but \textbf{SIMD vectorization within a single core}. Instruction sets such as AVX2 and AVX-512 can process 8--16 elements per instruction without introducing barriers, kernel wakeups, or coherency traffic, and therefore offer a more credible route to lower latency while preserving full core utilization.
